---
title: "Instructor reference: A molecule, a mouse, and a memory"
output:
  html_document:
    toc: true
    self_contained: true
    mathjax: null
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir=normalizePath(".."))
```

## Numerical reference, not a required conclusion

Students should interpret the observed estimates and their limits; do not assess whether
their hypothesis was supported. This calculation uses the same prepared animal table as
the course. The subtraction and interval are **Memantine minus Saline**.

```{r core-results}
d <- read.csv("data/study.csv")
first <- d$BDNF[d$group == "Saline"]
second <- d$BDNF[d$group == "Memantine"]
data.frame(group=c("Saline","Memantine"),n=c(length(first),length(second)),
 mean=c(mean(first),mean(second)),median=c(median(first),median(second)),
 SD=c(sd(first),sd(second)))
mean(second)-mean(first)
t.test(second,first)
wilcox.test(second,first,exact=FALSE)
```

## What to listen for

- One row represents one animal under the documented preparation rule; counting repeated cells or dilution readings inflates animal n.
- A larger SD describes more variation among animals. It is not the standard error of a mean.
- The outlier puzzle needs a fifth value of 34; mean 10.8, median 5. Four fixed readings imply the median stays 5 for every fifth value.
- Missingness exercise: observed n 3, observed mean 6, zero-substituted mean 4.5. Exclusion may still be biased.
- Histogram bins and violin smoothing change a display, not the measurements. Neither establishes population shape from a small sample.
- A Welch estimate targets means; the rank test is not generally a median comparison. Methods should not be chosen by the smallest p-value.
- A p-value does not give the probability that a hypothesis is true. Lack of significance does not demonstrate equivalence.
- Within-group association avoids one obvious pooling confound, but does not establish causation or remove shared assay effects.
- The invented Bonferroni example returns 0.05, 0.15, 1, 1, 0.20. Only the first meets the declared <= 0.05 threshold.

## Within-group association reference

```{r association-results}
for(g in c("Saline","Memantine")) {
 cat("Group:",g,"\n")
 s<-d[d$group==g,]
 print(cor.test(s$BDNF,s$pCREB,method="pearson"))
 print(cor.test(s$BDNF,s$pCREB,method="spearman",exact=FALSE))
}
```

Read `guide.md` and `data/dictionary.html` before teaching. Day-2 answers depend on the selected
question. The validation script runs every allowed combination; an exploratory result remains
exploratory even when p is small.
