---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

### Notebook 2. Data visualization

<br>

```{r}
# Let's re-import the data to start a new session
neurites <- read.csv(file="/Users/marek/SU drive/Teaching/2024/statistical analysis in R/Data/03_neurites_synthetic.csv")

# It's a good routine to display the first few rows to see if you data has imported properly
head(neurites)

```


#### Visualizing the Mean of Samples

Now, we will calculate and visualize the mean of neurite lengths for each condition using a barplot. Barplots are useful for comparing central tendencies across different groups.


```{r}
# First calculate means for all 3 conditions and store it in variables mean1, mean2 and mean3 
# Calculate means for the three conditions and store them in separate variables
mean_control <- mean(neurites$control)
mean_cond1 <- mean(neurites$condition_1)
mean_cond2 <- mean(neurites$condition_2)

# Create basic barplot
barplot(c(mean_control, mean_cond1, mean_cond2))


# Create a barplot with axis labels and color
barplot(c(mean_control, mean_cond1, mean_cond2), 
        names.arg = c('Control', 'Condition 1', 'Condition 2'),
        col = 'lightblue', 
        main = 'Mean Neurite Length by Condition',
        ylab = 'Mean Length [um]')

# Note: The c() function is used to concatenate the three mean values into a single vector for plotting.
```

#### Plotting Individual Observations

In addition to summarizing the data with means, it's helpful to look at the individual observations. We can plot each observation against its index (its row in the dataset).

```{r}
# Plot individual neurite measurements against their observation index
plot(unlist(neurites), 
     main = 'Individual Observations of Neurite Length', 
     xlab = 'Observation Index', 
     ylab = 'Neurite Length [um]', 
     col = 'darkblue', 
     pch = 19)

# Note: This plot shows all observations in a single plot. Since the data is sorted by condition, the plot remains organized.
# If the data were unsorted, this plot might appear disorganized.
```
#### Saving Plots to a PDF File

To preserve your plots for use in reports or presentations, R allows you to save them to various file formats, including PDFs.

```{r}
# Open a PDF file to save the plot
pdf(file = 'Neurites_plot.pdf', width = 6, height = 4)


# Create the same barplot inside the PDF file
barplot(c(mean_control, mean_cond1, mean_cond2), 
        names.arg = c('Control', 'Condition 1', 'Condition 2'),
        col = 'lightblue', 
        main = 'Mean Neurite Length by Condition', 
        ylab = 'Mean Length [um]')


# Close the PDF device to finish saving the plot
dev.off()

# Check the current working directory to confirm where the PDF was saved
getwd()

```

-----

#### Exercise: Open and examine the saved PDF file in your File Explorer.

-----

#### Visualizing the Data with a Boxplot

A boxplot provides a visual summary of the distribution of values, highlighting the median, quartiles, and potential outliers.


```{r}
# Create a boxplot to compare the distribution of neurite lengths across the three conditions
boxplot(neurites) 

# A nicer version of the same boxplot with colors, and axis labels
boxplot(neurites, 
        main = 'Boxplot of Neurite Lengths by Condition', 
        ylab = 'Neurite Length [um]', 
        col = c('lightblue', 'lightgreen', 'lightcoral'), 
        names = c('Control', 'Condition 1', 'Condition 2'))

# Note: Boxplots are useful for identifying the spread of the data, as well as potential outliers.

```

-----

#### Converting Data to Long Format

Wide format is useful when measurements are organized by separate variables (e.g., different conditions), but for certain analyses, it’s often necessary to reshape the data into long format.

```{r}
# Load the reshape2 library to convert the data
library(reshape2)

# Convert the dataset to long format
neurites_long <- melt(neurites)

# Display the first few rows of the long-format dataset
head(neurites_long)

# Check the dimensions of the dataset in long format
dim(neurites_long)

# Note: In long format, the 'variable' column contains the condition (Control, Condition 1, Condition 2),
# and the 'value' column contains the neurite length measurements.
```

#### Filtering Long Format Data

Once the data is in long format, we can filter for specific conditions more easily.

```{r}
# Filter only 'Control' observations
neurites_long[neurites_long$variable == 'control',]

# Filter only 'Condition 1' observations
neurites_long[neurites_long$variable == 'condition_1',]

# In fact what we are doing is asking which entry in the column variable of object neurites_long is equal to control/condition_1 and then using that vector as index to subsed the whole data frame 
neurites_long$variable == 'control'

```

#### Advanced Visualization with ggplot2

ggplot2 is a powerful package for creating advanced visualizations. Here, we will reproduce some of our previous plots using ggplot2 with enhanced aesthetics.

#### Basic barplot in ggplot2

```{r}
library(ggplot2)
library(ggridges)
library(ggthemes)

# ggplot always uses long format for plotting
# Basic bar plot using ggplot2
ggplot(data = neurites_long) + geom_bar(aes(x=variable,y=value), stat='summary',fun.y='mean')

# ggplot(data=) == initializes a ggplot object
# geom_bar adds a layer with barplot
# stat=summary tells it to use summary statistics and fun.y tells it to use mean as functon to calculate summary statistics
# aes() contains information about what should be on x and y axes

# Let's add some color to the plot by using fill=variable and black outline of the bars
ggplot(data = neurites_long) + geom_bar(aes(x=variable,y=value,fill=variable), stat='summary',fun.y='mean',col='black')

# We can also easily change the theme of the plot by adding theme layer - theme_base() makes it look more like a base R plot
ggplot(data = neurites_long) + geom_bar(aes(x=variable,y=value,fill=variable), stat='summary',fun.y='mean',col='black') + theme_base()

```

#### Histogram in ggplot2

```{r}
# Let's plot histogram in ggplot - this histogram ignores the information about the variable. It just pools our control, and conditions into a single graph
ggplot(data=neurites_long) + geom_histogram(aes(x=value))

# To split the histograms into 3 different ones, we can use facet_wrap layer to split the plots
ggplot(data=neurites_long) + geom_histogram(aes(x=value)) + facet_wrap(~variable) 

# Let's add some color again
ggplot(data=neurites_long) + geom_histogram(aes(x=value,fill=variable),col='black') + facet_wrap(~variable) # Creates 3 histograms for our 3 variables

# And change the theme
ggplot(data=neurites_long) + geom_histogram(aes(x=value,fill=variable),col='black') + facet_wrap(~variable) + theme_classic()


```

#### Bboxplot and jitterplot in ggplot

```{r}
# Basic boxplot
ggplot(data=neurites_long) + geom_boxplot(aes(x=variable,y=value))

# Add some color
ggplot(data=neurites_long) + geom_boxplot(aes(x=variable,y=value,fill=variable))

# Because of layer based syntax we can stack layers on top of each other
# Let's add individual datapoints to the boxplot
ggplot(data=neurites_long) + geom_boxplot(aes(x=variable,y=value,fill=variable)) + geom_point(aes(x=variable,y=value))

# The points are all in one row, we can add some jitter to avoid overplotting 
ggplot(data=neurites_long) + geom_boxplot(aes(x=variable,y=value,fill=variable)) + geom_jitter(aes(x=variable,y=value),width = 0.25)


# If we skip the boxplot, we can just have a jitter plot
ggplot(data=neurites_long) +  geom_jitter(aes(x=variable,y=value),width = 0.25)

# And with some colors
ggplot(data=neurites_long) +  geom_jitter(aes(x=variable,y=value,col=variable),width = 0.25)

# Let's now fix  some axis labels
ggplot(data=neurites_long) +  geom_jitter(aes(x=variable,y=value,col=variable),width = 0.25) + xlab("condition") + ylab("Neurite length [um]")
```

#### Density ridge histogram and violin plot

Density ridge plots and violin plots provide alternative ways to visualize the distribution of data across conditions.


```{r}
ggplot(data=neurites_long) + geom_density_ridges(aes(x=value,y=variable,fill=variable)) + theme_bw()
```


```{r}
# Violin plot
ggplot(data = neurites_long) + geom_violin(aes(y=value,x=variable,fill=variable)) + theme_bw()

# With dots
ggplot(data = neurites_long) + geom_violin(aes(y=value,x=variable,fill=variable)) + geom_jitter(aes(x=variable,y=value),width=0.15) + theme_bw() 
```


