---
title: "A molecule, a mouse, and a memory: my investigation"
output:
  html_document:
    toc: true
    self_contained: true
    mathjax: null
---

```{r setup, include=FALSE}
root <- normalizePath(if (file.exists("Statistics_lab.Rproj")) "." else "..")
stopifnot(file.exists(file.path(root,"Statistics_lab.Rproj")))
knitr::opts_knit$set(root.dir=root)
knitr::opts_chunk$set(echo=TRUE,fig.width=8,fig.height=4.5)
```

**Name:** write here  
**Partner:** write here

## Your report starts here

Today you choose the question and defend the analysis. Reuse code from yesterday; writing
functions from memory is not the goal. Each student submits **one integrated HTML notebook**
with code, output, captions and their own reasoning. Save this as `04_my_investigation.Rmd`.
Add useful material from your earlier notebooks where it helps explain your decisions.

Work in pairs and swap roles, but write your own interpretation. No prior statistical vocabulary
is assumed: use yesterday's notebook and ask for explanations whenever a term is unclear.

## 1. Choose a question before looking for a result — 20 minutes

| Question | Keep fixed | Compare |
|---|---|---|
| Does memantine change BDNF in trisomic mice? | genotype = Ts65Dn; learning = C/S | Saline vs Memantine |
| Does genotype relate to BDNF under saline? | treatment = Saline; learning = C/S | Control vs Ts65Dn |
| Does learning protocol relate to BDNF in control mice? | genotype = Control; treatment = Saline | S/C vs C/S |

Genotype describes control or trisomic mice. C/S means context then shock; S/C means shock
then context. These protocols differ in their opportunity for context learning. They are
not repeated measurements of the same animals. `BDNF`, `pCREB` and `NR1` are assay signals
in relative units, not absolute concentrations. They came from nuclear cortex fractions.
Do not equate an assay difference with improved memory or an effect in humans.


*My question: ___.*  
*I predict ___ because ___.*  
*My outcome, comparison direction and biological sample unit: ___.*

Choose one primary group comparison and one association. These are exploratory teaching
analyses of a published dataset, not a new confirmatory study. Do not try every option and
present only the smallest p-value.

## 2. Inspect and select — 25 minutes

```{r full-data}
full <- read.csv("data/investigation.csv")
head(full)
dim(full)
anyDuplicated(full$sample_id)
```

The selection scaffold below introduces `subset()` and `&` (both conditions must hold).
It keeps the experimental design explicit. `if` chooses the block matching your question.
You only need to change `question_choice`; explain which rows it retains before running it.

```{r choose-question}
question_choice <- 0 # Change to 1, 2 or 3 after writing your question.
outcome_name <- "BDNF" # Keep BDNF for the group comparison.
if (question_choice == 1) {
  selected <- subset(full, genotype == "Ts65Dn" & learning == "C/S")
  selected$group <- selected$treatment
  group_a <- "Saline"; group_b <- "Memantine"
} else if (question_choice == 2) {
  selected <- subset(full, treatment == "Saline" & learning == "C/S")
  selected$group <- selected$genotype
  group_a <- "Control"; group_b <- "Ts65Dn"
} else if (question_choice == 3) {
  selected <- subset(full, genotype == "Control" & treatment == "Saline")
  selected$group <- selected$learning
  group_a <- "S/C"; group_b <- "C/S"
} else {
  selected <- NULL
  cat("Choose one question above before continuing.")
}
unit <- "BDNF assay signal (relative units)"
```

`selected[[outcome_name]]` retrieves the column whose name is stored as text. We use it here
so the same worked code can support your chosen question. The earlier `$` notation is equally
useful when you type a column name directly.

```{r check-observations}
if (!is.null(selected)) {
  selected$value <- selected[[outcome_name]]
  table(selected$group)
  table(is.na(selected$value))
  complete <- selected[!is.na(selected$value), ]
  complete$group <- factor(complete$group, levels=c(group_a,group_b))
  table(complete$group)
  bdnf_by_group <- split(complete$value, complete$group)
  bdnf_by_group # Each vector is labelled with its biological group.
}
```

`split()` keeps a separate measurement vector for each named group in `bdnf_by_group`.
For example, `bdnf_by_group[[group_a]]` retrieves the group whose name is stored in
`group_a`. Displaying the collection shows the actual group names and measurements,
so these labels stay correct when you choose a different biological question.

Here `is.na()` marks missing values, `!` reverses those marks, and the bracket selection keeps
the observed values. You met missingness yesterday; this supplied pattern now automates the
counting and selection. It never replaces a missing value with zero.

*Rows included, exclusions, observed sample sizes and what one row means: write here.*

## 3. Give the reader a picture — 30 minutes

```{r inspect}
if (!is.null(selected)) {
  boxplot(value ~ group, data=complete, outline=FALSE,ylab=unit)
  set.seed(7)
  stripchart(value ~ group,data=complete,vertical=TRUE,method="jitter",add=TRUE,pch=19)
  par(mfrow=c(1,2))
  qqnorm(bdnf_by_group[[group_a]],main=group_a); qqline(bdnf_by_group[[group_a]])
  qqnorm(bdnf_by_group[[group_b]],main=group_b); qqline(bdnf_by_group[[group_b]])
  par(mfrow=c(1,1))
  mean(bdnf_by_group[[group_a]]); mean(bdnf_by_group[[group_b]])
  median(bdnf_by_group[[group_a]]); median(bdnf_by_group[[group_b]])
  sd(bdnf_by_group[[group_a]]); sd(bdnf_by_group[[group_b]])
  shapiro.test(bdnf_by_group[[group_a]]); shapiro.test(bdnf_by_group[[group_b]])
}
```

*Caption: measured quantity and units, groups, sample sizes, one visible pattern, one limitation.*

*Assumptions: discuss independence, the shapes, unusual observations and what the small
sample prevents you from knowing. Do not use Shapiro p > 0.05 as proof of normality.*

**15-minute break.**

## 4. Choose, run and interpret — 35 minutes

Write your primary method and reason **before** changing `method_choice` below. Welch compares
means; Wilcoxon uses ranks and is not generally a median test. Both need independent animals.
With very small samples neither is a magic fix for limited information.

*My primary method and its target: ___. My justification and main limitation: ___.*

```{r group-test}
method_choice <- "choose" # Change to "welch" or "wilcoxon".
if (!is.null(selected)) {
  mean_difference <- mean(bdnf_by_group[[group_b]]) - mean(bdnf_by_group[[group_a]])
  mean_difference # group_b minus group_a
  if (method_choice == "welch") {
    group_result <- t.test(bdnf_by_group[[group_b]],bdnf_by_group[[group_a]])
    print(group_result)
  } else if (method_choice == "wilcoxon") {
    group_result <- wilcox.test(bdnf_by_group[[group_b]],bdnf_by_group[[group_a]],exact=FALSE)
    print(group_result)
  } else cat("Choose and justify your method above.
")
}
```

*Result: state direction, observed magnitude, units, both sample sizes, method, p-value and
confidence interval if provided. A descriptive mean difference does not become the target of
a rank test. Explain what the result does and does not support.*

## 5. Investigate one association — 30 minutes

For your association, use the same selected subset and choose **one** of its groups.
Investigate BDNF versus pCREB (a phosphorylated signalling protein) within that group.
Do not combine all 72 mice and ignore their experimental conditions.

```{r correlation}
association_group <- "choose" # Use exactly one of your selected group labels.
correlation_method <- "choose" # "pearson" or "spearman", chosen after inspecting the plot.
if (!is.null(selected) && association_group %in% selected$group) {
  association <- selected[selected$group == association_group, ]
  x <- association$BDNF
  y <- association$pCREB
  keep <- complete.cases(x,y) # Keep animals with both measurements.
  x <- x[keep]; y <- y[keep]
  length(x)
  plot(x,y,pch=19,xlab="BDNF assay signal (relative units)",ylab="pCREB assay signal (relative units)")
  if (correlation_method == "pearson") print(cor.test(x,y,method="pearson"))
  if (correlation_method == "spearman") print(cor.test(x,y,method="spearman",exact=FALSE))
} else cat("Choose one group before running the association.
")
```

*Explain your scatterplot, coefficient, method, sample size and uncertainty. Pearson describes
linear association; Spearman describes rank association. What other process could explain the
relationship? Why does correlation alone not establish causation?*

**10-minute break.**

## 6. The tempting hunt for a small p-value — 20 minutes

Imagine ten unrelated comparisons when all ten null hypotheses are true. Each test conducted
at a 0.05 threshold has a 5% false-positive rate under its assumptions. Trying many tests raises
the chance of at least one false alarm. For independent tests it is `1 - 0.95^10`; real protein
measurements can be correlated, so that exact calculation need not apply to them.

These are **invented p-values**, not results from our animals:

```{r multiple-testing}
demo_p <- c(0.01,0.03,0.20,0.80,0.04)
adjusted_p <- p.adjust(demo_p,method="bonferroni")
data.frame(raw_p=demo_p,bonferroni_p=adjusted_p)
```

Bonferroni multiplies each p-value by the number of tests in the chosen family, capped at 1.
It controls the chance of any false rejection in that family under valid individual tests,
even when tests are dependent, though it can be conservative. Define the family from the
questions actually investigated, including unsuccessful attempts; not only the reported hits.
At a declared threshold of adjusted p <= 0.05, which example remains? What is lost by treating
0.049 and 0.051 as completely different scientific conclusions?

*My explanation and the family of questions I explored: write here.*

## 7. Peer review and a reproducible hand-in — 40 minutes

Give your partner three minutes to read your best figure and result. They ask:

1. Which animals contributed, and why are they independent?
2. Does the method answer the stated question?
3. Is the conclusion proportionate to the estimate, uncertainty and sample size?

*One useful challenge from my partner and how I responded: ___.*

Write a short final conclusion: answer your question, identify your strongest evidence,
state the most important limitation, and propose one useful next experiment.

If you used an LLM or another source: *I asked for ___; I checked ___; I now understand ___.*
You may ask for debugging or explanations; verify named columns, selected rows and scientific
claims. Keep only code and reasoning you can explain. Equivalent instructor help is available.

```{r completion-check, echo=FALSE}
ready <- !is.null(selected) && method_choice %in% c("welch","wilcoxon") &&
  association_group %in% selected$group && correlation_method %in% c("pearson","spearman")
cat(if(ready) "Analysis choices are filled. Review results and written explanations before submission."
    else "DRAFT: complete the question, test and association choices before submission.")
```

- [ ] The HTML includes my own question, data description, figures and captions.
- [ ] It reports a group comparison and a correlation with interpretation and limitations.
- [ ] I explained exclusions and analysis-specific sample sizes.
- [ ] I identified the questions explored and explained multiple testing.
- [ ] I restarted R, ran from the first chunk, knitted and checked the resulting HTML.
- [ ] All “write here” placeholders are replaced, and the draft notice is resolved.

Submit this integrated **HTML** to the learning system; keep the `.Rmd` as your editable recipe.

## Extra: 1,080 rows, but how many mice?

The source workbook contains 15 assay records per mouse: three replicates of a five-point
dilution series. The supplied teaching CSV retains record `_1` consistently for each mouse.
This sacrifices repeated-measurement information to keep the introductory analysis transparent.
It is not a claim that record 1 is the best estimate of true abundance. No outcome or p-value
was used to choose it. Explain why treating all 1,080 rows as independent mice would give
misleading precision, and why averaging across dilution levels needs assay knowledge.

**Design your follow-up:** What additional measurement or experimental design would distinguish
a change in an assay signal from a change in learning? Specify the animal-level sample size,
control condition and outcome before collecting data.


## Data source

[https://archive.ics.uci.edu/dataset/342/mice+protein+expression](https://archive.ics.uci.edu/dataset/342/mice+protein+expression). See `data/dictionary.html` for preparation rules and the original study.
