---
title: "04. Your MMSE investigation"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
    self_contained: true
    mathjax: null
---

[Course index](../index.html) · [Day 1 reference](03_statistical_analysis.html) · [Data dictionary](../data/dictionary.html)

# Day 2 — work in pairs, explain the analysis yourself

**Your name:** write here. **Working partner:** write here.

Today you will investigate a larger dementia dataset. Work with a partner: discuss questions, inspect plots, and help each other with code. Each of you should run the analysis in your own notebook and write your own explanations of what you did and why.

**Submit one integrated notebook per student**, with code, figures, and narrative. Save its `.Rmd` source and rendered HTML; these are two versions of the same notebook, not a separate report. Follow the instructor's submission instructions for uploading them. Instructors assess only your final submitted notebook. You do not need to have your plan, code, or results checked during the workshop.

You will first build a descriptive picture of the participants, then complete **one group comparison and one correlation**, plus a short response to the worked multiple-testing example. You do not need a large collection of tests. Write a few sentences beside each important result rather than leaving all the writing until the end.

## Today's plan

| Time after the start | Task |
|---|---|
| 0:00–0:15 | Recap and discuss the task |
| 0:15–0:35 | Worked example: multiple testing |
| 0:35–1:15 | Audit, explore participant characteristics, and write your plan |
| 1:15–1:30 | Break |
| 1:30–2:20 | Group comparison: plots, assumptions, test, and interpretation |
| 2:20–2:45 | One correlation and its scatterplot |
| 2:45–2:55 | Break |
| 2:55–3:05 | Challenge one conclusion |
| 3:05–3:40 | Complete your explanations |
| 3:40–4:00 | Restart, rerun, check, and discuss |

## Setup — run first

Save a personal copy in `notebooks/`, for example `04_mmse_investigation_Anna.Rmd`.

```{r setup, include=FALSE}
project_root <- normalizePath(if (file.exists("Statistics_lab.Rproj")) "." else "..",
                              mustWork = TRUE)
stopifnot(file.exists(file.path(project_root, "Statistics_lab.Rproj")))
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(echo = TRUE, error = FALSE, fig.width = 7, fig.height = 4.5)
```

## 1. Worked example: why account for multiple testing?

Imagine three planned treatment-versus-control comparisons in one experiment. These three questions form a **family of tests**. The p-values below are made up for this arithmetic example; they are not results from our dementia dataset.

If we try many tests and report only the smallest p-value, we hide the number of opportunities to find an apparently unusual result. Bonferroni adjustment accounts for how many tests we are considering together: multiply each p-value by the number of tests and cap the result at 1.

```{r bonferroni-example}
p_values <- c(Treatment_A = 0.01, Treatment_B = 0.03, Treatment_C = 0.20)
adjusted <- p.adjust(p_values, method = "bonferroni")
data.frame(comparison = names(p_values), raw_p = p_values,
           bonferroni_p = adjusted)
```

**Discuss, then answer:**

1. Which three comparisons are we considering together? Why do we multiply each p-value by three?
2. Which values are below 0.05 before and after adjustment?
3. Did the estimated treatment effects change, or did our criterion for evidence change?
4. Why would reporting only Treatment A's unadjusted p-value be incomplete?

**Your explanation:**

For your own investigation, write down two questions below and report both results, whether significant or not. You are exploring possible patterns, rather than confirming a discovery. Running two tests gives more opportunities for a small p-value than running one; mention this when discussing your results. Do not keep changing variables or methods until a small p-value appears. The optional adjustment at the end shows how to account for both tests when claiming that at least one has found a difference or relationship.

## 2. What are these data?

The observations come from the OASIS-2 longitudinal study. The previous course file contained **336 visits from 136 people**. This version keeps the **first visit of each of those 136 participants**, preserving the original course values and rounding. We restored the participant IDs from the original source.

The two labels, `Demented` and `Nondemented`, are the source study's longitudinal group classifications. Participants in its `Converted` group were already absent from the previous course file and remain excluded. This selection limits generalisation. These are observational data, not a randomised treatment experiment. Comparing MMSE across related clinical groups does not independently validate a diagnostic test.

```{r import}
mmse <- read.csv("data/mmse_baseline.csv")
head(mmse)
```

### Data dictionary

| Column | Meaning | How to treat it |
|---|---|---|
| `Subject_ID` | Source participant identifier | Identifier, not a measurement |
| `Source_row`, `Visit` | Original row reference and visit number | Provenance; all retained visits are 1 |
| `Group` | Source longitudinal group | Nominal category |
| `M.F` | Source-recorded M/F category | Categorical; do not infer unrecorded identity information |
| `Age` | Age at the visit, years | Numeric |
| `EDUC` | Years of education | Numeric, recorded in whole years |
| `SES` | Socioeconomic category: 1 highest status to 5 lowest | Ordinal; inspect category counts |
| `MMSE` | Cognitive assessment score, 0–30 | Bounded integer score |
| `CDR` | Clinical Dementia Rating | Ordered clinical categories |
| `eTIV` | Estimated intracranial volume, cm³ | Numeric; not a direct measure of functioning brain tissue |
| `nWBV` | Normalised whole-brain volume, stored as a fraction | Numeric; 0.70 corresponds to 70% on this scale |
| `ASF` | Atlas scaling factor | Unitless imaging quantity; not a primary outcome here |

Definitions and sources are in the [full data dictionary](../data/dictionary.html).

### Task A — inspect before analysing

Use the day 1 commands to answer:

1. How many rows and columns are there? Does each `Subject_ID` appear once?
2. How many participants are in each group?
3. Which variables contain missing values?
4. What are the observed age and MMSE ranges? Are they plausible given the dictionary?
5. Why should we avoid calculating a mean SES as if the categories were equally spaced measurements?

```{r student-audit}
# Add your checks here.
```

**Your data description:**

<details>
<summary>Hint 1: functions to reuse</summary>

Try `dim()`, `names()`, `table()`, `range()`, and `is.na()`. `unique()` keeps distinct values; compare the number of unique IDs with the number of rows.

</details>

<details>
<summary>Hint 2: code to adapt if you are stuck</summary>

This chunk is a reference. Copy needed lines into your audit chunk above so they run when you Knit.

```{r audit-hint, eval=FALSE}
dim(mmse)
length(unique(mmse$Subject_ID))
table(mmse$Group)
colSums(is.na(mmse))
range(mmse$Age, na.rm = TRUE)
```

`colSums(is.na(mmse))` counts missing entries in each column. Copy the last line and change the variable to inspect MMSE.

</details>

### Task A2 — who is in this sample? — 20 minutes

Build a short descriptive overview before the hypothesis tests. Use the neurite practice to help you select columns and adapt plot commands. Work through these four tasks; write a sentence beside each output. These are descriptions, not additional statistical tests.

1. Calculate the mean, median, SD, and range of age. Make an age histogram with a meaningful title and axis label. Try two bin widths or numbers of bins, and keep the view that best communicates the distribution.
2. Count participants by source-recorded M/F category using `table(mmse$M.F)`. Make a bar chart of the counts. Explain why this is a bar chart rather than a histogram.
3. Inspect the SES counts with `table(mmse$SES, useNA = "ifany")`. Report how many values are missing and why replacing them with zero would introduce a new, invalid category.
4. Select participants aged 80 or older. Count them and inspect their group counts. State the denominator when reporting their proportion of the full sample. This threshold is a descriptive exercise, not a new definition of disease.

```{r student-descriptive-overview}

```

**What kind of sample are we analysing? Include the age summary, category counts, and missingness in three or four sentences.**

<details>
<summary>Hints for the overview</summary>

Use `hist(mmse$Age, breaks = 8, xlab = "Age (years)", main = "Participant ages")` as a starting point. Use `barplot(table(mmse$M.F), xlab = "Source-recorded category", ylab = "Participants")` for counts. Adapt `mmse[mmse$Age >= 80, ]` to create a selected table; `nrow()` counts its participants. Age has no missing values here. For a variable with missing values, include an explicit missingness check in the selection.

</details>

## 3. Plan two questions before looking at test results

Our main group comparison is **MMSE between Demented and Nondemented participants**. What do you want to learn: **how much the average MMSE score differs between the groups**, or **whether scores tend to be higher in one group**? The first question compares means; the second uses the ordering of scores, called ranks. Start with the question you want to answer, then use the plots and checks below to justify your test. The supplied templates below use MMSE. If you choose another measurement, such as age, update the selected columns, plot labels, and interpretation throughout.

Choose **one** association: age with MMSE, **or** nWBV with MMSE. Choosing an association because its plot is relevant to your question is different from testing many associations and retaining only a significant one.

### Task B — write your plan

- **Which groups are you comparing, and what difference do you want to estimate?** For example: “Do Demented and Nondemented participants differ in average MMSE score? I will estimate Demented minus Nondemented in MMSE points.” Write your own question here:
- **Which two measurements will you compare, and what relationship do you want to investigate?** Choose age and MMSE, or nWBV and MMSE:
- **What I expect, and why:**
- **What does one row represent? Does any person appear more than once? Why does this matter for your analysis?**
- **Which values does a participant need for each analysis?** For example, can you include someone in the age–MMSE correlation if their age is missing?
- **Which plots will you make, and what will you check before choosing a test?** Think about the shape and spread of each group, extreme values, and whether measurements come from different people:

Read through your plan with your partner, then continue. You can change your initial choice of test after checking the data; explain why you changed it. Use the break after completing the plan.

## 4. Group comparison

The following chunk keeps participants with an observed group and MMSE score. It leaves unrelated missing values alone. Selection is based on the variables needed for this analysis, not on whether every column in a row is complete.

```{r prepare-groups}
group_data <- mmse[complete.cases(mmse[, c("Group", "MMSE")]), ]
group_data$Group <- factor(group_data$Group, levels = c("Nondemented", "Demented"))
nondemented <- group_data$MMSE[group_data$Group == "Nondemented"]
demented <- group_data$MMSE[group_data$Group == "Demented"]
table(group_data$Group)
nrow(mmse) - nrow(group_data)  # excluded from this comparison
```

### Task C — summarise and plot

Calculate mean, median, SD, and IQR for each group. Reuse notebook 1. Report sample sizes. A summary table is useful, but a list of clearly labelled results is sufficient.

```{r student-summaries}
# Your summary commands here.
```

Adapt the following plot labels if needed. The dots show individual observations; horizontal jitter separates tied values without changing their scores.

```{r group-plot}
boxplot(MMSE ~ Group, data = group_data, outline = FALSE,
        col = c("lightblue", "peachpuff"), ylim = c(0, 31),
        ylab = "MMSE score (points)", xlab = "Source group")
set.seed(7)
stripchart(MMSE ~ Group, data = group_data, method = "jitter",
           vertical = TRUE, add = TRUE, pch = 16, cex = 0.6, jitter = 0.15)
```

**Your caption, including group sizes, centre, spread, and any ceiling effect:**

### Task D — assess distributions

Use the supplied panel and normality tests. Read the plots first.

```{r group-distributions, fig.height=6}
par(mfrow = c(2, 2))
hist(nondemented, breaks = seq(-0.5, 30.5, 1), xlim = c(0, 31),
     ylim = c(0, 35),
     main = "Nondemented", xlab = "MMSE score (points)")
hist(demented, breaks = seq(-0.5, 30.5, 1), xlim = c(0, 31),
     ylim = c(0, 35),
     main = "Demented", xlab = "MMSE score (points)")
qqnorm(nondemented, main = "Nondemented: Q–Q plot")
qqline(nondemented)
qqnorm(demented, main = "Demented: Q–Q plot")
qqline(demented)
par(mfrow = c(1, 1))
```

Both histograms use the same axes and bins. If you analyse a different measurement, update these limits and labels as well as the selected values.

```{r shapiro-groups}
shapiro.test(nondemented)
shapiro.test(demented)
```

**What do your checks tell you?** Describe the shape of each group in the graphs and what its Shapiro result suggests. Do the score limit, repeated values, or extreme values make a normal distribution a questionable approximation? Explain why the Shapiro p-value alone cannot choose the test for you.

### Task E — choose your main test and explain why

Return to your question. Welch estimates a difference in means. Wilcoxon rank-sum uses ranks; it does not generally test a median difference without additional shape assumptions. Both require measurements from separate, independent participants; repeated visits from the same person would need a different approach. Recall the side-by-side comparison from day 1.

**Which test will I use, and why does it fit my question and these data? Write this before running the test:**

Enter either `"welch"` or `"wilcoxon"` between the empty quotes below. The supplied code runs only that method. You do not need to learn `if` statements today; they select the appropriate supplied command.

```{r chosen-group-test}
group_method <- ""
if (group_method == "welch") {
  group_result <- t.test(demented, nondemented, var.equal = FALSE)
  print(mean(demented) - mean(nondemented))
  print(group_result)
} else if (group_method == "wilcoxon") {
  group_result <- wilcox.test(demented, nondemented, exact = FALSE)
  print(group_result)
} else {
  cat("Choose welch or wilcoxon in group_method after writing your justification.\n")
}
```

### Task F — explain the result

Include group sample sizes, direction, relevant summaries, method, statistic, and p-value. For Welch, include the mean difference and its 95% CI in the direction **Demented minus Nondemented**. For Wilcoxon, report the rank-based result and descriptive median/IQR values; do not claim its p-value tests the observed difference in medians or invent a mean-difference CI.

**Your result paragraph:**

**What this result does not establish:**

## 5. One correlation

### Task G — choose the variables and inspect the scatterplot

Enter `"Age"` or `"nWBV"` below, following your plan. Leave the selection fixed when comparing plain and group-coloured displays.

```{r correlation-plot}
correlation_variable <- ""
if (correlation_variable %in% c("Age", "nWBV")) {
  correlation_data <- mmse[complete.cases(mmse[, c(correlation_variable, "MMSE")]), ]
  x <- correlation_data[[correlation_variable]]
  y <- correlation_data$MMSE
  x_label <- if (correlation_variable == "Age") "Age (years)" else "Normalised whole-brain volume (fraction)"
  plot(x, y, pch = 16, xlab = x_label, ylab = "MMSE score (points)")
  print(nrow(correlation_data))
  print(nrow(mmse) - nrow(correlation_data))
} else {
  cat("Choose Age or nWBV in correlation_variable to draw your scatterplot.\n")
}
```

`mmse[[correlation_variable]]` selects the column whose name is stored in that object. For example, when the name is `"Age"`, it selects the same values as `mmse$Age`. `%in%` checks whether your choice is one of the two allowed names.

**Describe your scatterplot:** As one measurement increases, does the other tend to increase, decrease, or show no clear pattern? Does the pattern look roughly straight or curved? Are there clusters or isolated points that might strongly affect the result? How many participants are shown?

### Task H — select a correlation method

Use `"pearson"` to ask how closely the two measurements follow a straight-line pattern (**linear association**). Use `"spearman"` to ask whether higher values of one measurement tend to go with higher or lower values of the other, using their ranks. This trend can be curved, but should keep the same overall direction (**monotonic association**). Explain your choice using the plot, your question, and the MMSE score scale. Do not choose solely from separate Shapiro tests of the two variables.

**My choice and reason, before running the test:**

```{r chosen-correlation}
correlation_method <- ""
if (correlation_variable %in% c("Age", "nWBV") &&
    correlation_method %in% c("pearson", "spearman")) {
  correlation_result <- cor.test(x, y, method = correlation_method, exact = FALSE)
  print(correlation_result)
} else {
  cat("Choose your variable, run its plot chunk, then choose pearson or spearman.\n")
}
```

For Spearman we request an approximate p-value because some participants have the same values (called ties). For Pearson, the usual p-value and confidence interval assume independent pairs of measurements and an appropriate model for their joint distribution. A clear curve or a few extreme points should make you question this analysis; R returning a number does not settle whether the method fits. Base R does not return a Spearman confidence interval.

**Your result:** Give the correlation value (r or rho), say whether the relationship is positive or negative, and report the number of participants and p-value. Explain one thing the result cannot tell us. Include the confidence interval if using Pearson. Explain why a coefficient near zero would not rule out every possible relationship.

## 6. Challenge one conclusion — ten minutes

Choose one of these questions. You do not need another statistical test.

1. Does colouring your chosen scatterplot by group change your interpretation of the relationship seen when both groups are shown together?
2. How does the ceiling of the MMSE scale affect your group plot and summaries?
3. Why would using all 336 visits as independent people be inappropriate?

If you choose question 1, copy and run this reference chunk after your correlation plot:

```{r group-colour-hint, eval=FALSE}
point_colours <- ifelse(correlation_data$Group == "Demented", "#B34D22", "#246B91")
plot(x, y, col = point_colours, pch = 16, xlab = x_label,
     ylab = "MMSE score (points)")
legend("bottomleft", legend = c("Demented", "Nondemented"),
       col = c("#B34D22", "#246B91"), pch = 16, bty = "n")
```

**Your answer and how it changes or qualifies your conclusion:**

## 7. Using LLMs and other help

Feel free to ask an LLM for debugging help, hints, or a more detailed explanation of a concept. Include the actual column names, your question, and the error or command you are working on. Ask for a small step if you are lost.

Useful requests include:

- “Explain this error and suggest the smallest correction. Here is my code and the exact message.”
- “Explain why Welch and Wilcoxon ask different questions using a small numerical example. Then ask me a question to check my understanding.”
- “Help me interpret this R output. Which conclusion would go beyond what it shows?”

Check that suggested code uses the intended variables, groups, and sample sizes. Compare the explanation with the actual output. You should be able to explain any command you retain. Do not accept a method simply because an LLM recommended it, and write your own narrative.

**If you used an LLM:** Give one short example: what you asked, what you checked, and what you now understand. A chat transcript is unnecessary. If you did not use one, say so; instructor help and the hints are equally acceptable.

## 8. Finish your notebook

You have 35 minutes to finish explanations and 20 minutes to rerun, check, and discuss. Use the work already written beside each output.

- [ ] I explained what each row represents, which groups are included, and which values are missing.
- [ ] I completed the age, M/F, SES, and age-selection overview with plots and explanations.
- [ ] I stated one comparison and one correlation question before testing.
- [ ] I described the plots and explained why I chose my main group test.
- [ ] I reported results with sample sizes, relevant estimates, and limitations.
- [ ] I included a scatterplot and explained the correlation.
- [ ] I answered the separate multiple-testing questions.
- [ ] I challenged one conclusion and described assistance if used.
- [ ] I wrote my own explanations and replaced the empty answer prompts.
- [ ] I restarted R and knitted the notebook successfully.

Make sure the rendered notebook contains actual group-test and correlation results, not the messages asking you to choose a method. Save both versions of your notebook. On a room computer, copy your files to a location you can access after the class.

**Final conclusion, three or four sentences:** Answer your two questions, state the main limitations, and identify one thing these data cannot establish. A non-significant result is still a result worth explaining.

## Extra: a four-panel figure — 15 minutes

Once the required analysis and writing are complete, bring your findings together in a figure: age histogram, MMSE boxplot by group, your chosen scatterplot, and SES category counts. Reuse the commands you already understand.

Start with `par(mfrow = c(2, 2))`, add your four plotting commands, and finish with `par(mfrow = c(1, 1))`. For the SES panel, the following commands label the missing category explicitly. Copy them into your four-panel chunk as its fourth plot. Add titles and axis labels to every panel. Do not add more hypothesis tests.

```{r ses-panel-reference, eval=FALSE}
ses_counts <- table(mmse$SES, useNA = "ifany")
ses_labels <- names(ses_counts)
ses_labels[is.na(ses_labels)] <- "Missing"
barplot(ses_counts, names.arg = ses_labels, main = "SES including missing values",
        xlab = "SES category", ylab = "Participants")
```

```{r student-four-panel, fig.width=9, fig.height=7}

```

**Caption:** Name the sample, describe one finding from each panel, and state a limitation. What does the combined figure make easier to understand? Which details would need a larger separate plot?

## Extra: explore another measurement — 10 minutes

Choose **EDUC or nWBV**. Before running anything, use the dictionary to state its meaning and units. Make a histogram, calculate appropriate summaries, and describe the values by group using a boxplot. Write one question the plots raise that your two planned analyses cannot answer. Do not turn it into another test today.

```{r student-extra-measurement}

```

**Your observation and unanswered question:**

## Extra: adjustment for your two planned tests

This is optional. If your conclusion is “at least one of my two tests found a difference or relationship”, account for the fact that you had two opportunities to obtain a small p-value. After completing both analyses, copy this chunk into your notebook and run it. It does not justify running extra tests until you obtain significance.

```{r optional-adjustment, eval=FALSE}
my_p_values <- c(group = group_result$p.value, correlation = correlation_result$p.value)
data.frame(raw_p = my_p_values,
           bonferroni_p = p.adjust(my_p_values, method = "bonferroni"))
```

[R reference for multiple-testing adjustment](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/p.adjust.html)
