---
title: "Day 1 report · MMSE teaching comparison"
author: "Marek Bartosovic"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

<!-- This is YOUR report. Write sentences outside the grey chunks. Code goes inside them.
     Run a chunk with the green triangle. Knit (button above) to produce day1_report.html.
     Replace every line that starts with "WRITE HERE". -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## 1. Question and data

WRITE HERE: one sentence stating the question we ask today.

```{r import}
mmse <- read.csv("data/mmse_small.csv")
head(mmse)
dim(mmse)
table(mmse$group)
control <- mmse$MMSE[mmse$group == "Control"]
ad <- mmse$MMSE[mmse$group == "AD"]
```

WRITE HERE: what does one row represent, and what kind of variable is each column?

## 2. Summaries

```{r summaries}
tapply(mmse$MMSE, mmse$group, mean)
tapply(mmse$MMSE, mmse$group, median)
tapply(mmse$MMSE, mmse$group, sd)
tapply(mmse$MMSE, mmse$group, IQR)
sum(is.na(mmse$MMSE))
```

WRITE HERE: which centre and spread would you report for each group, and why? Are any values missing?

## 3. Figure

```{r figure, fig.width=6, fig.height=4.5}
boxplot(MMSE ~ group, data = mmse, ylim = c(0, 30),
        xlab = "Teaching group", ylab = "MMSE (points)", col = c("#9bd5db", "#f4bb8b"))
set.seed(1)
stripchart(MMSE ~ group, data = mmse, vertical = TRUE,
           method = "jitter", add = TRUE, pch = 16)
```

WRITE HERE: a two-sentence caption, with one observed pattern and one limitation.

## 4. Method and assumptions

```{r assumptions, fig.width=7, fig.height=3.5}
par(mfrow = c(1, 2))
qqnorm(control, main = "Control MMSE"); qqline(control)
qqnorm(ad, main = "AD MMSE"); qqline(ad)
shapiro.test(control)
shapiro.test(ad)
```

WRITE HERE, using the worksheet: I inspected ___; I observed ___; I chose ___ because ___; an important limitation is ___.

## 5. Result

```{r tests}
t.test(ad, control)
wilcox.test(ad, control, exact = FALSE)
```

| | Welch's t-test | Wilcoxon rank-sum |
|---|---|---|
| Question it answers | WRITE HERE | WRITE HERE |
| Main assumptions | WRITE HERE | WRITE HERE |
| Result (estimate / statistic, p) | WRITE HERE | WRITE HERE |
| One limitation | WRITE HERE | WRITE HERE |

WRITE HERE: the result sentence. "In this teaching sample (n = … per group), the estimated AD minus Control difference in mean MMSE was … points (95% CI … to …, Welch's t-test, p …). This supports … under the assumption that …; it does not show …."

## 6. Association (worked example)

```{r association, fig.width=5, fig.height=4}
set.seed(7001)
signal_a <- seq(1, 10, length.out = 30)
signal_b <- 2 + 0.7 * signal_a + rnorm(30, sd = 1.8)
plot(signal_a, signal_b, pch = 16,
     xlab = "Assay A (arbitrary units)", ylab = "Assay B (arbitrary units)")
cor.test(signal_a, signal_b, method = "pearson")
```

WRITE HERE: direction and form of the association, the coefficient with n, and one limitation.

## 7. Help I used and how I checked it

WRITE HERE: "I asked (partner / instructor / LLM) for help with ___; I checked the suggestion by ___; I now understand ___." One example is enough.

## 8. Reproducibility check

Restart R (Session → Restart R), then Knit. If the HTML appears without errors, the analysis is reproducible.

```{r session}
sessionInfo()$R.version$version.string
```
