---
title: "A molecule, a mouse, and a memory: R basics and visualisation"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
source("helpers.R", local=TRUE)
knitr::opts_chunk$set(echo=FALSE,fig.width=8,fig.height=4.5)
tutorial_options(exercise.checker=checker,exercise.error.check.code="# Friendly feedback",exercise.timelimit=15,exercise.lines=8)
```

## Welcome: you are the analysis team

BDNF is a protein involved in neuronal plasticity. Researchers measured proteins in the nuclear fraction of mouse cortex after a learning protocol. Our first comparison uses control-genotype mice given saline or memantine. A change in one assay signal does not establish improved memory.

**Your question:** Does memantine change the BDNF assay signal in control mice exposed to the same learning protocol?

No previous R or statistics is needed. First we will learn to run a command, then follow the
measurements into a figure. You are allowed to be surprised by the data. A result that does
not support your prediction is still a result worth explaining.

Work in pairs. One person types and the other predicts or explains; swap after each section.
Keep your own written answers. Allow 70 minutes for basics, a break, 35 for plots, 35 for the biological-replication investigation and 5 for the handoff.

- **Code editor:** type R commands in the box containing code. Click **Run Code** to see results below it.
- **Answer box:** type your scientific explanation in ordinary words.
- **Submit Answer:** save your current code attempt and request feedback. Run Code alone does not save an attempt for download.

Keep the supplied commands and **append your solution underneath**, unless a task explicitly
asks you to repair an error. Each editor runs independently: keep its supplied import lines.
`#` starts a comment, which R ignores. You can experiment without changing the source data files.

```{r identity}
textInput("student","Your name")
textInput("partner","Partner")
downloadButton("notebook","Download my notebook (.Rmd)")
```

Download regularly. Written answers are session-only; download before closing or reloading.
Your notebook contains your submitted code, your explanations and your current slider views.
It will become an ordinary editable R notebook in RStudio.

## 1. First successful runs — 20 minutes

Click inside the editor and press Enter for a new line. Running an editor runs every line,
top to bottom. Output appears in the same order. The printed `[1]` marks the position of the
first output item; it is not part of the calculated number.

### Make R do the arithmetic

Run the supplied addition. Then append `12 / 4` on a new line, run again and submit. `/` means divide.

```{r divide, exercise=TRUE}
2 + 3

# Add your command below.
```

```{r divide-check}
# Checked by checker.
```

```{r divide-hint-1}
# Keep the example and append 12 / 4.
```

```{r divide-solution}
2 + 3
12 / 4
```

### Plan the plate

Three plates, four wells each: how many wells? Predict first. Append `3 * 4`; `*` means multiply.

```{r multiply, exercise=TRUE}
3 + 4

# Add your command below.
```

```{r multiply-check}
# Checked by checker.
```

```{r multiply-hint-1}
# Append 3 * 4 below the example.
```

```{r multiply-solution}
3 + 4
3 * 4
```

### Give a result a name

An **object** is information kept under a name. Read `wells <- 12` as “store 12 in wells”. The arrow is `<` followed by `-`. A name on its own displays its value. Keep the example, then append `wells <- 16` and `wells` on separate lines.

```{r store, exercise=TRUE}
wells <- 12
wells
```

```{r store-check}
# Checked by checker.
```

```{r store-hint-1}
# Append wells <- 16, then wells.
```

```{r store-solution}
wells <- 12
wells
wells <- 16
wells
```

### Keep several measurements together: c()

`c()` combines values into a **vector**, an ordered collection. A **function** is a named operation; its inputs go inside parentheses. In `c(2, 4, 6)`, commas separate three values. Nothing is being added yet! Run the example. Then append `counts <- c(2, 4, 6, 8)` and `counts` to keep and display four values.

```{r combine, exercise=TRUE}
counts <- c(2, 4, 6)
counts
```

```{r combine-check}
# Checked by checker.
```

```{r combine-hint-1}
# Append counts <- c(2, 4, 6, 8), then counts.
```

```{r combine-solution}
counts <- c(2, 4, 6)
counts
counts <- c(2, 4, 6, 8)
counts
```

### An average in two ways: mean()

The **mean** is the sum divided by the number of observations. The example calculates it by hand. Append `mean(counts)`: give the stored vector to the function by putting its name inside parentheses. Does it agree? `length(counts)` reports how many values there are, not their average.

```{r average, exercise=TRUE}
counts <- c(2, 4, 6)
(2 + 4 + 6) / 3
length(counts)
```

```{r average-check}
# Checked by checker.
```

```{r average-hint-1}
# Append mean(counts) at the bottom.
```

```{r average-solution}
counts <- c(2, 4, 6)
(2 + 4 + 6) / 3
length(counts)
mean(counts)
```

### Predict before you press Run

These are made-up assay readings. Before running, predict whether adding 9 will raise or lower the mean. Keep the example and append a new vector containing 4, 5, 6, 9, then calculate its mean. Explain the change to your partner.

```{r add_value, exercise=TRUE}
readings <- c(4, 5, 6)
mean(readings)
```

```{r add_value-check}
# Checked by checker.
```

```{r add_value-hint-1}
# Append readings <- c(4, 5, 6, 9), then mean(readings).
```

```{r add_value-solution}
readings <- c(4, 5, 6)
mean(readings)
readings <- c(4, 5, 6, 9)
mean(readings)
```

### Recover from a small error

Run this editor once. R names are case-sensitive: `Readings` and `readings` are different. Repair the capital letter and rerun. In RStudio, a Console prompt of `+` means an unfinished command; press Esc, then repair and rerun the complete command in your notebook.

```{r rescue, exercise=TRUE}
readings <- c(4, 5, 6)
mean(Readings)
```

```{r rescue-check}
# Checked by checker.
```

```{r rescue-hint-1}
# Use readings with a lower-case r in both lines.
```

```{r rescue-solution}
readings <- c(4, 5, 6)
mean(readings)
```

## 2. Meet the data — 10 minutes

One row is one mouse, using assay record 1 from that mouse. The original study measured each sample repeatedly; those repeated readings are not extra mice.

Our prepared table has **19 animals**: 9 Saline and 10 Memantine.
The original measurements are real, openly available research data. The preparation and exact source
are in `data/dictionary.html`. [Dataset](https://archive.ics.uci.edu/dataset/342/mice+protein+expression) · [Original study](https://doi.org/10.1371/journal.pone.0129126).

| Column | Meaning |
|---|---|
| `sample_id` | Animal identifier; a label, not a measured quantity |
| `group` | Saline or Memantine |
| `BDNF` | BDNF assay signal (relative units) |
| `pCREB` | pCREB assay signal (relative units), from the same animal |

`BDNF` is the brain-derived neurotrophic factor assay signal. `pCREB` is the signal for phosphorylated CREB, a protein involved in regulating gene activity. Both are relative assay signals from the same mouse.

A table in R is called a **data frame**. A row holds information about one animal;
a column holds one variable. A **nominal** category names groups with no ranking;
an **ordinal** category has an order but need not have equal gaps. A **continuous** measurement
can take intermediate values even if the instrument rounds them. Biological meaning matters,
not just whether R stores digits. A day number and a named learning phase are different descriptions.

### Open the table

`read.csv()` reads a comma-separated table. Quotes mark literal text, including a file path. `<-` stores the table as study. `head()` previews rows, `names()` lists columns, and `dim()` reports rows then columns. Run the supplied import, then append `head(study)`, `names(study)` and `dim(study)` in that order.

```{r import, exercise=TRUE}
study <- read.csv("data/study.csv")
```

```{r import-check}
# Checked by checker.
```

```{r import-hint-1}
# Append head(study), names(study), then dim(study).
```

```{r import-solution}
study <- read.csv("data/study.csv")
head(study)
names(study)
dim(study)
```

```{r note-import}
note("row")
```

## 3. Follow one group — 10 minutes

`study$BDNF` retrieves the measurement column. `study$group == "Saline"` asks a yes/no
question for each row. **Two equals signs compare; `<-` assigns.** Square brackets keep
the values whose group comparison is TRUE. The selection code is supplied: you need to
explain and adapt it, not memorise it.

The name `saline_bdnf` tells us both the biological measurement and its group. We will use `memantine_bdnf` for the other group. Descriptive names help you check which animals a calculation includes.

### Follow the Memantine animals

Run the Saline example. Append a selection for `"Memantine"`, store it as `memantine_bdnf`, then append `length(memantine_bdnf)` and `mean(memantine_bdnf)`. Say which animals contributed before reading the mean.

```{r select, exercise=TRUE}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
length(saline_bdnf)
mean(saline_bdnf)
```

```{r select-check}
# Checked by checker.
```

```{r select-hint-1}
# Append memantine_bdnf <- study$BDNF[study$group == "Memantine"], then length(memantine_bdnf) and mean(memantine_bdnf).
```

```{r select-solution}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
length(saline_bdnf)
mean(saline_bdnf)
memantine_bdnf <- study$BDNF[study$group == "Memantine"]
length(memantine_bdnf)
mean(memantine_bdnf)
```

```{r note-select}
note("selection")
```

## 4. Typical values, variation and an awkward reading — 20 minutes

The mean gives each observation equal weight. The **median** is the middle of the ordered
observations (the mean of the two middle ones when there is an even number).
The **sample standard deviation (SD)** describes variation among observations, in the same
units as the measurement. A larger SD means more variation; it does not by itself tell us
how uncertain the mean is. Sample size also matters for that.

### Which group varies more?

Run the supplied mean and median calculations. Append `sd_saline_bdnf <- sd(saline_bdnf)` and `sd_memantine_bdnf <- sd(memantine_bdnf)`, then display each named result on its own line. Compare the SDs in your answer.

```{r spread, exercise=TRUE}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
memantine_bdnf <- study$BDNF[study$group == "Memantine"]
mean(saline_bdnf)
median(saline_bdnf)
mean(memantine_bdnf)
median(memantine_bdnf)
```

```{r spread-check}
# Checked by checker.
```

```{r spread-hint-1}
# Append sd_saline_bdnf <- sd(saline_bdnf) and sd_memantine_bdnf <- sd(memantine_bdnf); display both.
```

```{r spread-solution}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
memantine_bdnf <- study$BDNF[study$group == "Memantine"]
mean(saline_bdnf)
median(saline_bdnf)
mean(memantine_bdnf)
median(memantine_bdnf)
sd_saline_bdnf <- sd(saline_bdnf)
sd_memantine_bdnf <- sd(memantine_bdnf)
sd_saline_bdnf
sd_memantine_bdnf
```

```{r note-spread}
note("spread")
```

### Challenge: double the headline

A colleague reports that “the average response doubled”. Four of five made-up culture
readings are fixed at 4, 5, 5 and 6. Only the fifth reading can change. Can you double
the original mean without changing the median? Work out a candidate before moving the slider.
These are separate invented readings, not edits to our animal measurements.

```{r slider-outlier}
sliderInput("extreme","Fifth assay reading (arbitrary units)",0,40,7,step=1)
plotOutput("assay_plot")
verbatimTextOutput("assay_stats")
```

### Test the headline in code

Keep the original readings. Append a new assay vector with your chosen fifth value, then `new_mean <- mean(assay)` and `new_median <- median(assay)`. Display both results. Bonus: with the other four readings fixed, which median values are even possible?

```{r outlier, exercise=TRUE}
assay <- c(4, 5, 5, 6, 7)
mean(assay)
median(assay)
```

```{r outlier-check}
# Checked by checker.
```

```{r outlier-hint-1}
# The original sum is 27. Double it to 54; subtract the four fixed values. Store both new summaries.
```

```{r outlier-solution}
assay <- c(4, 5, 5, 6, 7)
mean(assay)
median(assay)
assay <- c(4, 5, 5, 6, 34)
new_mean <- mean(assay)
new_median <- median(assay)
new_mean
new_median
```

```{r note-outlier}
note("outlier")
```

## 5. The assay that failed — 10 minutes

`NA` means missing. It does not mean zero. Imagine one assay failed: would writing zero
describe what happened? First predict what R should do when a required input is unknown.
Then compare two decisions using these invented values; keep the source table untouched.

### Missing is a different claim from zero

Run the starter: `mean(example)` returns NA. Count the recorded numbers by looking at the vector. Append `observed_n <- 3` and `observed_mean <- mean(example, na.rm = TRUE)`. The comma introduces another argument; `na.rm = TRUE` excludes missing entries for this calculation only. In a separate vector called `zero_example`, use 0 instead of NA; calculate `zero_mean <- mean(zero_example)`. Display the count and both means.

```{r missing, exercise=TRUE}
example <- c(4, 6, NA, 8)
mean(example)
observed_n <- length(example) # Change: count observed values, excluding NAs
```

```{r missing-check}
# Checked by checker.
```

```{r missing-hint-1}
# Count the three recorded numbers. Use mean(example, na.rm = TRUE); calculate the zero substitution separately.
```

```{r missing-solution}
example <- c(4, 6, NA, 8)
mean(example)
observed_n <- length(example) # Counts every slot, including NA.
observed_n <- 3
observed_mean <- mean(example, na.rm = TRUE)
zero_example <- c(4, 6, 0, 8)
zero_mean <- mean(zero_example)
observed_n
observed_mean
zero_mean
```

```{r note-missing}
note("missing")
```

Excluding a missing value does not recover it. If high values fail more often, the observed
values may systematically miss the high end of the distribution.

```{r checkpoint}
downloadButton("notebook_checkpoint","Download before the break")
```

**Take a 15-minute break.** On return, swap typing and explanation roles.

## 6. How many bars tell the story? — 15 minutes

A **histogram** groups measurements into intervals called bins. Bar height counts observations
inside each interval. A few broad bins can hide structure; many narrow bins can make a small
sample look full of gaps. Neither setting adds information. The marks below the bars show
where the actual observations lie. Both groups below use the same bins and axes.

Try to make the data look as smooth as possible, then as irregular as possible. Ask your
partner which biological conclusions survive both views. This is a challenge to the display,
not an invitation to select whichever plot supports your preferred story.

```{r histogram-controls}
radioButtons("hist_mode","Control the bins using",c("Number of bins"="count","Bin width"="width"),inline=TRUE)
sliderInput("hist_bins","Number of bins",3,20,6,step=1)
```

```{r histogram-width}
sliderInput("hist_width","Bin width (measurement units)",0.00606836,0.182051,0.0303418,step=0.00182051)
plotOutput("hist_explorer")
textOutput("bin_summary")
```

### Make a histogram you can rerun

`hist()` draws a histogram. `breaks = 5` requests roughly five bins; base R chooses convenient boundaries, so the actual count can differ. The slider above uses exact boundaries. `main` sets the title and `xlab` labels the horizontal axis. Keep the Saline example and append `hist_memantine_bdnf <- hist(memantine_bdnf, breaks = 5, main = "Memantine", xlab = "BDNF assay signal (relative units)")`.

```{r histogram, exercise=TRUE}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
memantine_bdnf <- study$BDNF[study$group == "Memantine"]
hist(saline_bdnf, breaks=5, main="Saline", xlab="BDNF assay signal (relative units)")
```

```{r histogram-check}
# Checked by checker.
```

```{r histogram-hint-1}
# Keep the saline_bdnf plot and append hist_memantine_bdnf <- hist(memantine_bdnf, breaks=5, main="Memantine", xlab="Measurement units").
```

```{r histogram-solution}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
memantine_bdnf <- study$BDNF[study$group == "Memantine"]
hist(saline_bdnf, breaks=5, main="Saline", xlab="BDNF assay signal (relative units)")
hist_memantine_bdnf <- hist(memantine_bdnf, breaks=5, main="Memantine", xlab="BDNF assay signal (relative units)")
```

```{r note-histogram}
note("histogram")
```

## 7. Show the animals behind the average — 10 minutes

`factor()` below fixes the display order so colours and group positions stay consistent across figures. It does not change any measurements.

A **boxplot** puts the middle half of the observations inside a box. The line is the median.
Whiskers extend to observations within 1.5 box lengths of the box; more distant observations
are conventionally drawn separately. These points deserve investigation, not automatic deletion.
With small samples, always look at the actual observations as well.

In R's plot formula `BDNF ~ group`, read `~` as “shown by group”. `data = study` tells R where
those column names live. `stripchart()` adds individual observations. Jitter moves points
sideways slightly to reveal overlap; it never changes their measured height. `set.seed(7)`
makes that sideways movement repeatable.

### Can the mean-only chart tell the whole story?

Run the supplied bar chart, which shows only two means. Append the following worked template to add a boxplot with all animals. Run, inspect and submit. You do not need to invent the plotting syntax.

```r
study$group <- factor(study$group, levels=c("Saline","Memantine"))
comparison <- boxplot(BDNF ~ group, data=study, outline=FALSE,
  ylab="BDNF assay signal (relative units)", xlab="Group", col=c("lightblue","peachpuff"))
set.seed(7)
stripchart(BDNF ~ group, data=study, vertical=TRUE, method="jitter",
  jitter=0.1, add=TRUE, pch=16)
```

```{r boxplot, exercise=TRUE}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
memantine_bdnf <- study$BDNF[study$group == "Memantine"]
barplot(c(mean(saline_bdnf),mean(memantine_bdnf)),names.arg=c("Saline","Memantine"),
  ylab="Mean: BDNF assay signal (relative units)",main="Only the means")
```

```{r boxplot-check}
# Checked by checker.
```

```{r boxplot-hint-1}
# Append the supplied boxplot and stripchart template below the mean-only example.
```

```{r boxplot-solution}
study <- read.csv("data/study.csv")
saline_bdnf <- study$BDNF[study$group == "Saline"]
memantine_bdnf <- study$BDNF[study$group == "Memantine"]
barplot(c(mean(saline_bdnf),mean(memantine_bdnf)),names.arg=c("Saline","Memantine"),ylab="Mean: BDNF assay signal (relative units)")
study$group <- factor(study$group, levels=c("Saline","Memantine"))
comparison <- boxplot(BDNF ~ group, data=study, outline=FALSE,
  ylab="BDNF assay signal (relative units)", xlab="Group", col=c("lightblue","peachpuff"))
set.seed(7)
stripchart(BDNF ~ group, data=study, vertical=TRUE, method="jitter",
  jitter=0.1, add=TRUE, pch=16)
```

```{r note-boxplot}
note("boxplot")
```

### A smooth violin from a small sample?

A violin's width estimates where measurements are more concentrated. The dots are observed;
the smooth outline is calculated. Each violin here has equal maximum width and stops at its
observed range; width does not encode sample size or a confidence interval. The short line is
the median. Change the smoothing: can you make a bump vanish without changing a single animal?

```{r violin-controls}
sliderInput("smooth","Smoothing multiplier",0.4,2.5,1,step=0.1)
plotOutput("violin")
note("violin")
```

## 8. Two measurements from the same animal — 10 minutes

A **scatterplot** gives each animal one point. Its horizontal and vertical positions show
two different measurements from that animal. A rising pattern suggests positive association;
a falling pattern suggests negative association. First inspect a single group, so a difference
between groups is not mistaken for a relationship within a group.

### Does the pattern travel to the other group?

In `study[condition, ]`, the part before the comma selects rows and the blank after it keeps all columns. Run the supplied Saline plot. Append the same three commands for Memantine, naming the selected table `memantine_mice`. Keep the same measurement axes and label the group.

```{r scatter, exercise=TRUE}
study <- read.csv("data/study.csv")
saline_mice <- study[study$group == "Saline", ]
plot(saline_mice$BDNF, saline_mice$pCREB,
  xlab="BDNF assay signal (relative units)", ylab="pCREB assay signal (relative units)", main="Saline", pch=19)
```

```{r scatter-check}
# Checked by checker.
```

```{r scatter-hint-1}
# Append memantine_mice <- study[study$group == "Memantine", ], then plot(memantine_mice$BDNF, memantine_mice$pCREB, xlab="Measurement", ylab="pCREB assay signal (relative units)").
```

```{r scatter-solution}
study <- read.csv("data/study.csv")
saline_mice <- study[study$group == "Saline", ]
plot(saline_mice$BDNF, saline_mice$pCREB,
  xlab="BDNF assay signal (relative units)", ylab="pCREB assay signal (relative units)", main="Saline", pch=19)
memantine_mice <- study[study$group == "Memantine", ]
plot(memantine_mice$BDNF, memantine_mice$pCREB,
  xlab="BDNF assay signal (relative units)", ylab="pCREB assay signal (relative units)", main="Memantine", pch=19)
```

```{r note-scatter}
note("scatter")
```

## 9. More rows, more evidence? — 35 minutes

### Two lab members, two sample sizes

One colleague says: “We have 1,080 observations.” Another says: “We have 72 mice.”
They are looking at the same experiment. **Can they both be right?** Write your prediction
on paper or tell your partner before opening the larger table.

So far we have used one row per animal. Now open `replicates.csv`, where a row is
one assay reading. It includes more experimental groups than our first comparison.
Do not choose a statistical test yet. First find out what was actually replicated.

### Count measurements, then count animals

`nrow()` counts table rows. `unique()` keeps each distinct ID once; it does not average or remove measurements from the source. Work in short steps: append `measurement_n <- nrow(replicates)`, then `animal_ids <- unique(replicates$sample_id)`, then `animal_n <- length(animal_ids)`. Display both counts on separate lines.

```{r count_replicates, exercise=TRUE}
replicates <- read.csv("data/replicates.csv")
head(replicates)
```

```{r count_replicates-check}
# Checked by checker.
```

```{r count_replicates-hint-1}
# Count rows with nrow(); store distinct sample_id values with unique(); count those IDs with length().
```

```{r count_replicates-solution}
replicates <- read.csv("data/replicates.csv")
head(replicates)
measurement_n <- nrow(replicates)
animal_ids <- unique(replicates$sample_id)
animal_n <- length(animal_ids)
measurement_n
animal_n
```

```{r note-count_replicates}
note("replication_count")
```

### Reveal: what was repeated?

The mouse study measured each sample at five dilution levels, with three assay replicates per level: 15 readings per mouse. These are repeated assay measurements of the same biological source. Their spread can reflect both dilution conditions and technical variation. **Do not average the 15 rows blindly** or treat their record numbers as time or as known dilution labels. Our introductory table deliberately selected record 1 consistently; it did not create new animals.

A **biological replicate** is an independent biological unit relevant to the question.
The **experimental unit** is the unit independently assigned to an experimental condition.
Repeated readings of the same sample are **technical replicates**; multiple cells within an
animal are biological subsamples, not simply repeated instrument readings. These distinctions
come from the experimental design, not from how many spreadsheet rows appear.

For our animal-level question, the animal is the biological unit we retain. A unique ID alone
does not prove independence: shared litters, cages, treatments or batches may matter. If a
treatment were assigned to whole cages, for example, counting every animal as an independent
treatment replicate could still be wrong.

### Follow mouse 309

The supplied selection keeps every reading belonging to one animal. Run it, then append `one_animal_n <- nrow(one_animal)` and display the count. Append this supplied plot to inspect the readings. `record_number` is a record index, **not time**.

```r
plot(one_animal$record_number, one_animal$BDNF, pch=19,
     xlab="Record within one animal (not time)", ylab="BDNF assay signal (relative units)")
```

```{r follow_animal, exercise=TRUE}
replicates <- read.csv("data/replicates.csv")
head(replicates)
one_animal <- replicates[replicates$sample_id == "309", ]
one_animal
```

```{r follow_animal-check}
# Checked by checker.
```

```{r follow_animal-hint-1}
# Keep the supplied selection, then append one_animal_n <- nrow(one_animal) and one_animal_n. Add the supplied plot.
```

```{r follow_animal-solution}
replicates <- read.csv("data/replicates.csv")
head(replicates)
one_animal <- replicates[replicates$sample_id == "309", ]
one_animal
one_animal_n <- nrow(one_animal)
one_animal_n
plot(one_animal$record_number, one_animal$BDNF, pch=19,
     xlab="Record within one animal (not time)", ylab="BDNF assay signal (relative units)")
```

```{r note-follow_animal}
note("replication_animal")
```

### Give the dots their owners

The next display uses a small selection of animals **from the same condition** so the plot
stays readable. First look at the pooled cloud. Then reveal animal IDs. The vertical positions
and the number of measurements stay fixed; the revealed groups show which observations share
an animal. Compare variation within an animal with differences between animals. This small
view illustrates the structure; it does not establish the structure of the entire population.

```{r replicate-controls}
checkboxInput("replicate_owners","Reveal which animal owns each dot",FALSE)
plotOutput("replicate_plot")
```

### The copy-and-paste experiment: can we manufacture precision?

Predict what will happen if we copy every row of the original comparison's reference group
five times. Will the mean change? Will the number of animals change? Could a computer be
fooled by the larger row count?

This is a **deliberately incorrect analysis of duplicated rows**, using the small animal-level
table from earlier. No new assays, cells or animals are collected. The source files stay intact.
This exaggerated example exposes an assumption; real repeated measurements need not be identical.

The **standard error (SE)** describes uncertainty in an estimated mean under a statistical model;
it is different from the SD describing spread among observations. A familiar SE calculation,
`SD / sqrt(n)`, assumes independent observations. The solid line below deliberately breaks that
assumption by pretending copied rows are independent. The dashed reference uses the original
animal-level values and stays fixed. It too relies on the original animals being independent.

```{r copy-controls}
sliderInput("copy_factor","Copies of each existing row (1 means original table)",1,10,1,step=1)
tableOutput("copy_counts")
plotOutput("copy_precision")
note("replication_copies")
```

This mistake is called **pseudoreplication**: treating dependent observations as if they independently
replicated the biological comparison. Extra readings can help estimate measurement variability
or characterise an animal better. They cannot substitute for collecting additional independent
biological replicates. A sophisticated statistical model must still respect that structure.

### Return to our actual biological question

The larger file contains other experimental groups. We must select the original comparison
before reporting its sample size. The filtering code is supplied; `&` means both conditions
must hold. The extra readings below must lead back to the same animals we started with.

### Defend the n in your report

Run the supplied selection and read the measurement counts. Append `n_saline <- length(saline_ids)` and `n_memantine <- length(memantine_ids)`, then display both. Explain why the animal counts—not the number of assay or cell rows—belong in our original animal-level comparison.

```{r report_animals, exercise=TRUE}
replicates <- read.csv("data/replicates.csv")
comparison_readings <- replicates[replicates$genotype == "Control" & replicates$learning == "C/S", ]
table(comparison_readings$treatment) # Measurement counts, not animal counts.
saline_ids <- unique(comparison_readings$sample_id[comparison_readings$treatment == "Saline"])
memantine_ids <- unique(comparison_readings$sample_id[comparison_readings$treatment == "Memantine"])
```

```{r report_animals-check}
# Checked by checker.
```

```{r report_animals-hint-1}
# Append n_saline <- length(saline_ids) and n_memantine <- length(memantine_ids), then display both counts.
```

```{r report_animals-solution}
replicates <- read.csv("data/replicates.csv")
comparison_readings <- replicates[replicates$genotype == "Control" & replicates$learning == "C/S", ]
table(comparison_readings$treatment) # Measurement counts, not animal counts.
saline_ids <- unique(comparison_readings$sample_id[comparison_readings$treatment == "Saline"])
memantine_ids <- unique(comparison_readings$sample_id[comparison_readings$treatment == "Memantine"])
n_saline <- length(saline_ids)
n_memantine <- length(memantine_ids)
n_saline
n_memantine
```

```{r note-report_animals}
note("replication_report")
```

**Your final recommendation to the lab:** “Our comparison contains ___ and ___ biological
units. Additional readings tell us ___, but do not independently replicate ___. Before choosing
an analysis I would check ___.” Exchange recommendations with your partner.

For these assays, a defensible summary must account for dilution and technical measurement structure; the fixed record-1 teaching subset is a transparent simplification, not a claim that this is the optimal assay analysis. A hierarchical model is an extension, not a command you need to learn today.

**Bridge to inference:** before asking whether two means differ, ask what independently
replicates the comparison. A smaller p-value cannot repair the wrong sample unit.

## 10. Your own notebook — 5 minutes

```{r reflection}
note("exit")
note("help")
textOutput("completion")
downloadButton("notebook_final","Download my completed basics notebook (.Rmd)")
```

1. Submit after your final edits, then download. Save the file in the project's `notebooks/` folder.
2. Open that Rmd in RStudio. **Source editor:** saved code and written explanations. **Console:** temporary execution and messages. **Environment:** temporary objects created by running code.
3. In Session → Restart R, clear those temporary objects. Run the setup chunk with its green triangle, then run the remaining chunks in order. Ctrl+Enter (Windows/Linux) or Cmd+Enter (Mac) runs the current line; the chunk's green triangle runs the whole chunk.
4. Type your prose outside code chunks. Keep the opening and closing chunk fences. Click **Knit** to make an HTML notebook. Knit starts a fresh R session; it cannot use objects created only in your Console.
5. Check the rendered HTML for errors, missing explanations and unfinished tasks. A successful Knit is not proof of a complete analysis.

Continue with `notebooks/03_statistical_analysis.Rmd` in RStudio. Day 2 uses
`notebooks/04_investigation.Rmd`. Keep your Rmd source and submit the final HTML to the learning system.
On a shared computer, save your download and use **Start Over** before the next student.

## Extra: the middle half

Quartiles split ordered data into four parts. The interquartile range, IQR, is the upper
quartile minus the lower quartile. Try `quantile(saline_bdnf)`, `IQR(saline_bdnf)` and `range(saline_bdnf)` in
your downloaded notebook after creating `saline_bdnf`. Explain why the range is especially sensitive
to a single extreme value. With small samples, software conventions for quartiles can differ.

```{r server, context="server"}
output$assay_plot <- renderPlot({
 x<-c(4,5,5,6,input$extreme)
 plot(x,1:5,xlim=c(0,42),pch=19,xlab="Made-up assay reading (arbitrary units)",ylab="Observation")
 abline(v=mean(x),col="#a54821",lwd=2); abline(v=median(x),col="#16776d",lwd=2,lty=2)
 legend("topright",c("Mean","Median"),col=c("#a54821","#16776d"),lty=c(1,2),lwd=2,bty="n")
})
output$assay_stats <- renderPrint({x<-c(4,5,5,6,input$extreme); c(mean=mean(x),median=median(x))})
output$hist_explorer <- renderPlot(hist_view(read.csv("data/study.csv"),input$hist_mode,input$hist_bins,input$hist_width))
output$bin_summary <- renderText({e<-hist_edges(read.csv("data/study.csv"),input$hist_mode,input$hist_bins,input$hist_width); paste(length(e)-1,"shared bins; width",signif(e[2]-e[1],3),"measurement units. Both axes are shared.")})
output$violin <- renderPlot(violin_view(read.csv("data/study.csv"),input$smooth))
output$replicate_plot <- renderPlot(replication_plot(read.csv("data/replicates.csv"),input$replicate_owners))
output$copy_counts <- renderTable(copy_summary(read.csv("data/study.csv"),input$copy_factor),digits=4)
output$copy_precision <- renderPlot(copy_plot(read.csv("data/study.csv"),input$copy_factor))
output$completion <- renderText({s<-learnr::get_tutorial_state(); passed<-sum(vapply(names(exercise_info),function(n)isTRUE(s[[n]]$correct),logical(1))); paste(passed,"of",length(exercise_info),"code checks passed. Review your explanations and rerun in RStudio too.")})
handler <- function() downloadHandler(filename=function()"01_my_basics_and_plots.Rmd",content=function(file){
 s<-isolate(learnr::get_tutorial_state())
 notes<-setNames(lapply(names(note_labels),function(n)input[[paste0("note_",n)]]),names(note_labels))
 settings<-list(mode=input$hist_mode,bins=input$hist_bins,width=input$hist_width,smooth=input$smooth,owners=input$replicate_owners,copies=input$copy_factor)
 writeLines(export_notebook(s,notes,input$student,input$partner,settings),file,useBytes=TRUE)
},contentType="text/plain; charset=utf-8")
output$notebook <- handler()
output$notebook_checkpoint <- handler()
output$notebook_final <- handler()
```
